SIMD\slash SIMT-vectorized Sparse Chemical Kinetic Jacobian\slash Thermo-chemical Source-Term Evaluation

### 2018-03-21 paper.tex
first pass at dr. sung edits

%====================================================================
\begin{abstract} % not to exceed 200 words
A code generation platform, \texttt{pyJac}, for single-instruction, multiple-data (SIMD) and single-instruction, multiple thread (SIMT) vectorized constant-pressure\slash constant-volume thermo-chemical source-term and sparse\slash dense chemical kinetic Jacobian evaluation was studied for a selection of chemical kinetic models covering the range of sizes typically used in reactive-flow simulations.
A new formulation of the chemical kinetic governing equations was derived and validated, resulting Jacobian sparsities of \SIrange{28.6}{92.0}{$\percent$} for the tested models.
Speedups of \SIrange{3.40}{4.08}{$\times$} were found for shallow-vectorized OpenCL source-rate evaluation compared a parallel OpenMP code on an \avx/ central processing unit (CPU), increasing to \SIrange{6.63}{9.44}{$\times$} and \SIrange{3.03}{4.23}{$\times$} for sparse and dense chemical kinetic Jacobian evaluation; in addition, the performance of a NVIDIA \gpunew/ graphics processing unit (GPU) was compared to a NVIDIA \gpuold/ GPU finding speedups of \SIrange{1.40}{1.88}{$\times$}, \SIrange{1.10}{1.59}{$\times$}, and \SIrange{1.36}{3.0}{$\times$} for source-rate, sparse and dense Jacobian evaluation, respectively.
Additionally, the effect of data-ordering was investigated and an storage pattern specifically formulated for vectorized evaluation proposed; as well, the effect of the \conp/\slash\conv/ assumptions and varying vector widths were studied on source-term evaluation performance.
Speedups over a first-order finite-difference approach reached \SIrange{17.60}{45.13}{$\times$} and \SIrange{55.11}{245.63}{$\times$} for sparse\slash dense evaluation on the CPU\slash GPU respectively, while dense Jacobian evaluation was up to \SI{19.56}{$\times$} and \SI{2.84}{$\times$} times faster than a previous version of \texttt{pyJac} on a \sse/ CPU and \gpuold/ GPU respectively.
Finally, future directions for vectorized chemical kinetic evaluation and sparse linear-algebra techniques were discussed.
\end{abstract}

@@

As the importance of detailed chemical kinetics for predictive reactive-flow simulations has become recognized~\cite{LU2009192}, the size and complexity of chemical kinetic models describing current and next-generation fuels relevant to transportation and power generation have increased greatly; e.g., a recent biodiesel model~\cite{WESTBROOK2011742} consists of \textasciitilde\num{3500} chemical species, and over \num{17000} reactions.
Moreover, the cost of evaluating the chemical source-terms scales linearly with the size of the model, while evaluation and factorization of the chemical kinetic Jacobian---if naively implemented via finite-difference method---scale quadratically and cubically with the number of species in the model~\cite{LU2009192}.
These factors often make the use of detailed chemical kinetics prohibitive in practice; e.g., in a recent direct numerical simulation~\cite{Spafford:2010aa} utilizing a 22-species model, evaluation of reaction rates consumed around half of the total run time.
In addition, chemical kinetic Jacobian evaluation and factorization---utilized by most common implicit integration techniques to deal with numerical stiffness---is a bottleneck for even modestly-sized chemical models in realistic reactive flow simulations, necessitating use of other cost-reduction strategies~\cite{LU2009192}.

A host of techniques have been developed to lessen the computational demand of chemical kinetic calculations while maintaining fidelity, these fall broadly into a few categories: removal of unimportant species and reactions unimportant species and reactions~\cite{Lu:2006bb,Pepiot-Desjardins:2008,Hiremath:2010jw,Niemeyer:2010bt,Curtis:2015}, lumping of species with similar thermo-chemical properties~\cite{Lu:2007,Ahmed:2007fa,Pepiot:2008kq}, and time-scale methods that reduce numerical stiffness~\cite{Maas:1992ws,Lam:1994ws,Lu:2001ve,Gou:2010}.
The interested reader is referred to the recent review of Tur\'anyi and Tomlin~\cite{turanyi2016analysis} for a comprehensive overview.

@@

SIMD and SIMT programming are two important vector-processing paradigms used increasingly in scientific computing.
Traditional multi-core parallelism is often used to increase central processing unit (CPU) performance, however as exponential growth in processing power---colloquially known as Moore's law---has slowed~\cite{khan2018science}, SIMD processors as well as SIMT processors, e.g., in the form of graphics processing units (GPUs), have gained recognition due to their increased floating operation throughput.
The parallel programming standard OpenCL~\cite{stone2010opencl} has further enabled adoption of vector-processing based codes in scientific computing by providing a common application program interface (API) for execution on heterogeneous systems, e.g., the CPU, GPU, or Intel's Many Integrated Core architecture (MIC), etc.
In this discussion, we will largely use OpenCL terminology to describe these processing paradigms, as it provides a convenient way to classify otherwise disparate processor type (e.g., CPUs and GPUs), however the concepts discussed are broadly applicable to SIMD\slash SIMT processing and not tied to OpenCL specifically.

\begin{figure}[htb]
  \centering
  \begin{subfigure}[t]{0.45\linewidth}
      \includegraphics[width=\textwidth]{SIMD.pdf}
      \caption{Schematic of SIMD processing.  A single compute unit (e.g., a CPU core) contains a vector unit with $N_v$ processing elements, together called a vector-lane.  The vector unit executes a single instruction concurrently on multiple data.}
      \label{F:SIMD}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.45\linewidth}
      \includegraphics[width=\textwidth]{SIMT.pdf}
      \caption{Schematic of SIMT processing. A single compute unit (e.g., a GPU streaming multiprocessor) contains many processing elements and hosts many threads, each with an instruction to execute (I1, I2).  Threads with the same instruction execute concurrently on multiple data while the others must wait (leading to thread divergence).}
      \label{F:SIMT}
  \end{subfigure}
\end{figure}

A typical modern CPU has many compute units (i.e., cores), each with specialized vector processing units capable of running SIMD instructions (\cref{F:SIMD}).
A SIMD instruction utilizes the vector processor to execute the same floating point operation (e.g. multiplication, division, etc.) on a several different data concurrently; the number of concurrent operations possible is known as the vector-width\footnote{OpenCL allows for use of vector-widths different from the actual hardware vector-width via implicit conversion, and may provide some performance benefit as will be studied in Sec.~\ref{S:results}} and is typically around 2--4 double precision operations. 
Additionally, specialized hardware accelerators, e.g. Intel's Xeon Phi co-processor (MIC) have been developed that have tens of cores with very wide vector-widths (e.g. 4--8 double precision operations); these very wide vector-widths are also found on cutting-edge and forthcoming Intel CPUs; i.e., the Skylake Xeon and Cannon Lake architectures, respectively.

SIMT processing---the foundation of modern GPUs~\cite{lindholm2008nvidia}---is a related computing paradigm that hosts large numbers of threads on a single compute element (a streaming multiprocessor in NVIDIA terminology).
A group of threads---typically \num{32}, known as a warp on NVIDIA GPUs---execute the same SIMD instruction on multiple data concurrently (\cref{F:SIMT}).
If some threads must execute a different instruction---e.g., due to if\slash then branching or predication---they are forced to wait and execute later.
This phenomena, known as thread-divergence, is a key consideration for SIMT-processing and can cause serious performance degradation for complicated algorithms~\cite{CurtisGPU:2017}.

@@

Recognizing the need to accelerate Jacobian evaluation and factorization, a number of recent works have been published on analytical chemical kinetic Jacobian evaluation; although as will be discussed at the end of this section, this work offers several key improvements over past efforts.
The \texttt{TChem} software package~\cite{Safta:2011vn} was one of the first packages developed providing this functionality, but has several limitations including: incompatibilities with modern reaction types---i.e., pressure dependent Arrhenius (or P-Log) and Chebyshev reactions---and lack of thread-safety to enable parallel execution~\cite{Curtis2017:tchem}. 
Youssefi~\cite{Youssefi:2011tm} explored the importance of analytical Jacobian matrices for time-scale analysis techniques as well as their effect on computational efficiency in zero-dimensional homogeneous reactor simulations.
Bisetti~\cite{Bisetti:2012jw} developed an isothermal, isobaric analytical Jacobian code-generation utility; although incompatible with many combustion codes, a significant increase in Jacobian sparsity was exhibited.
Additionally, in the same work Bisetti provided a novel way to compute dense matrix-vector multiplications resulting from a change of system variables without need for storage of the full dense Jacobian.
Perini et al.~\cite{Perini:2012gy} developed an analytical Jacobian code for constant-volume combustion, with additional options to increase sparsity (at expect of strict correctness), and enable tabulation of temperature dependent properties; a \SI{80}{\percent} speedup over a finite-difference-based Jacobian was reported when used in a multidimensional reactive-flow simulation.
Dijkmans et al.~\cite{Dijkmans:2014bb} developed a GPU-based analytical Jacobian code with optional tabulation of temperature-dependent properties, and found speedups up to \SI{120}{$\times$} for zero-dimensional chemical kinetic integration with large chemical models (\textasciitilde\num{3000} species) with a Tesla C2075 GPU.
Niemeyer et al.~\cite{Niemeyer:2016aa} created and validated the open-source analytical chemical kinetic Jacobian code-generator, \texttt{pyJac}, supporting parallel execution on the CPU and SIMT execution on the GPU; a speedup of \SIrange{3}{7.5}{$\times$} over a finite-difference Jacobian was found on the CPU.
Gao et al.~\cite{GAO2015287} derived a sparse analytical Jacobian, but it was not validated outside the context of use with an implicit-integration technique; in addition the Jacobian was based on an over-constrained system, the effect on strict conservation of mass\slash energy was not studied.

@@

Niemeyer and Sung~\cite{Niemeyer:2014aa} later developed a stabilized explicit second-order Runge--Kutta--Chebyshev algorithm on a Tesla C2075, demonstrating an order of magnitude speedup for a over a CPU implementation of VODE on a six-core Intel X5650 for moderately stiff chemical kinetics.

@@

In evaluation of the thermo-chemical properties, speedups between \SIrange{8}{13}{$\times$} were found (over a serial version of the same code) while running on 16 CPU cores, and \SIrange{20}{40}{$\times$} on the GPU; speedups of \textasciitilde\SI{9}{$\times$} and \textasciitilde\SI{25}{$\times$} were found in solution of a partially premixed methanol flame PDE on 16 CPU cores and the GPU respectively.
Curtis et al.~\cite{CurtisGPU:2017} implemented a fifth-order implicit Runge--Kutta method~\cite{wanner1991solving}, as well as two-fourth order exponential integration techniques~\cite{Hochbruck:1998,Hockbruck:2009} paired with an analytical Jacobian code~\cite{Niemeyer:2016aa} on a Tesla C2075; the implicit Runge--Kutta method was roughly equivalent to a standard implicit integrator~\cite{Hindmarsh:2005} running on 12--38 Intel Xeon E5-4640 v2 CPU cores for two smaller chemical models with an integration time step of \SI{e-6}{$\sec$}.

In contrast, SIMD-based chemical kinetics evaluation\slash integration have been far less studied.
Linford et al.~\cite{Linford:2011} implemented a three-stage, second-order Rosenbrock integrator for atmospheric chemical kinetics on the CPU, GPU and cell croadband engine (CBE)---a specially designed vector processor---and found speedups regularly exceeding~\SI{25}{$\times$} over a serial CPU implementation.
Kroshko and Spiteri~\cite{kroshko2013efficient} implemented a SIMD-vectorized third-order stiff Rosenbrock integrator for atmospheric chemistry on the CBE and found a speedup of \SI{1.89}{$\times$} (a parallel scaling efficiency of \SI{0.94}{$\percent$}) over a serial version of the same code.
Stone et al.,~\cite{stone2016} implemented a linearly-implicit fourth-order stiff Rosenbrock solver in the OpenCL for various platforms including CPUs, GPUs, and MICs.
SIMD-vectorization improved the integrator performance over an OpenMP baseline (which was vectorized by simple compiler hints (i.e., \texttt{\#pragmas}) by \SIrange{2.5}{2.8}{$\times$} on the CPU and \SIrange{4.7}{4.9}{$\times$} on the MIC, while the GPU performance was only \SIrange{1.4}{1.6}{$\times$} faster than the OpenMP baseline due to thread-divergence concerns.

This work will build upon our previous analytical chemical kinetic Jacobian code, \texttt{pyJac}~\cite{Niemeyer:2016aa}, in order to:
\begin{itemize}
 \item Derive and validate a new Jacobian formulation for \texttt{pyJac} that greatly increases sparsity
 \item Enable cross-platform SIMD\slash SIMT vectorization for the CPU, GPU and other accelerators
 \item Investigate the performance for a wide range chemical kinetic models, specifically looking at speedups due to SIMD\slash SIMT-vectorization as well as relative to the previous version of \texttt{pyJac}, and finally
 \item Discuss future extensions to this work as well as the most promising directions for SIMD\slash SIMT vectorization in reactive-flow simulations
\end{itemize}
To our knowledge, this is the first effort that enables vectorized chemical-kinetic source-term and Jacobian evaluation for any chemical model on a wide-selection of platforms, i.e., the CPU, GPU and other accelerators.

@@

\begin{figure}[htb]
  \centering
  \begin{minipage}{0.45\linewidth}
    \begin{subfigure}[t]{\textwidth}
      \includegraphics[width=\textwidth]{data_layouts.pdf}
      \caption{A simple 2-D data array with K rows and N columns.}
      \label{F:mem}
    \end{subfigure}
  \end{minipage}
  \hfil
  \begin{minipage}{0.45\linewidth}
    \begin{subfigure}[t]{\textwidth}
	\includegraphics[width=\textwidth]{row_major.pdf}
	\caption{Row-major data ordering}
	\label{F:row_major}
    \end{subfigure}
    \\
    \begin{subfigure}[t]{\textwidth}
	\includegraphics[width=\textwidth]{column_major.pdf}
	\caption{Column-major data ordering}
	\label{F:column_major}
    \end{subfigure}
  \end{minipage}
\end{figure}

When storing arrays for a chemical kinetic model, the data-storage layout and vectorization patterns are critical to achieving high-performance code.
\Cref{F:mem} depicts an example data array with $K$-rows and $N$-columns, and index ($i$, $j$) corresponding to the $i$th row and $j$th column.
For example the concentration of species $j$ for the $i$th thermo-chemical state would be stored in $[C]_{i, j}$ with $1 \le i \le N_{state}$, the number of thermo-chemical states considered for evaluation, and $1 \le j \le \ns/$, the number of species in the model.
The stored concentrations would then have $K = N_{\text{state}}$ rows, and $N = N_{\text{sp}}$ columns.

In the ``C'' (C row-major) format, the concentrations of all species for a single thermo-chemical condition $i$ would be stored sequentially in memory, i.e., with  $[C]_{1, 1}$ in index \num{1}, $[C]_{1, 2}$ in index \num{2} etc., as shown in \cref{F:row_major}.
Conversely in the ``F'' (Fortran column-major) format, the concentrations of a single species $j$ over all thermo-chemical states are adjacent in memory, corresponding to storing $[C]_{1, 1}$ in index \num{1}, $[C]_{2, 1}$ in index \num{2} and so on, as shown in \cref{F:column_major}.
This ordering---along with the device (CPU, GPU, etc.\@) and vectorization pattern in question---have a large effect on the performance of SIMD\slash SIMT-vectorized algorithms.

@@

In a \textit{deep}-SIMD\slash SIMT vectorization (also referred to as ``per-block'' in previous GPU works), a compute-unit utilizes its SIMD-lanes\slash SIMT-threads cooperatively to evaluate the thermo-chemical source-terms for a single thermo-chemical state; thus SIMD-lanes loading $[C]_{1, j} \ldots [C]_{1, j + N_v}$---the species concentrations for state 1, species $j \ldots j + N_v$---will access sequential memory locations if the data is stored in ``C''-order.
Further, in ``C''-ordering the maximum distance between any two species concentrations within the same thermo-chemical state is $N_{\text{sp}}$, with $N_{\text{sp}} \ll N_{\text{state}}$ in most cases; this greatly improved data-locality increases the chances of a cache-hit on the CPU, but may lead to uncoalesced memory-accesses on the GPU.
In addition, a deep vectorization requires both synchronization between SIMD-lanes\slash SIMT-threads via memory-fences\slash barriers (an expensive operation), and may result in SIMD-waste\slash SIMT thread-divergence, caused by different SIMD-lanes\slash SIMT-threads executing different instructions (e.g., resulting from different if\slash then branches).
It is noted that generally speaking, shallow-vectorizations may also experience SIMD-waste\slash SIMT thread-divergence, e.g., in chemical kinetic integration due to varying internal solver time-step sizes~\cite{CurtisGPU:2017}; however, in this work shallow-vectorizations are largely unaffected by this concern.

\begin{figure}[htb]
  \centering
  \begin{minipage}{0.6\linewidth}
    \begin{subfigure}[t]{\textwidth}
	\includegraphics[width=\textwidth]{row_major_split.pdf}
	\caption{Row-major, shallow-vectorized data ordering}
	\label{F:row_major_split}
    \end{subfigure}
    \\
    \begin{subfigure}[t]{\textwidth}
	\includegraphics[width=\textwidth]{column_major_split.pdf}
	\caption{Column-major, deep-vectorized data ordering}
	\label{F:column_major_split}
    \end{subfigure}
  \end{minipage}
  \caption{Vectorized data ordering patterns}
  \label{F:vector_data}
\end{figure}

@@

\subsection{Thermo-chemical Source-Terms and Jacobian}
This new version of \texttt{pyJac} is capable of evaluating the thermo-chemical source-terms for using both the constant-pressure (\conp/) or constant-volume (\conv/) assumption\footnote{Note: in this context, the ``constant-pressure'' and ``constant-volume'' assumptions refer to evaluation within a reaction sub-step in the operator splitting scheme, rather than a general constant-pressure or constant-volume reactive-flow simulation.}
In this section, we will only outline a brief summary of the system evaluated by \texttt{pyJac}; the interested reader is referred to the complete derivations, attached as supplemental material to this document.

>>>>

%====================================================================
\begin{abstract} % not to exceed 200 words
A code generation platform, \texttt{pyJac}, for single-instruction, multiple-data (SIMD) and single-instruction, multiple thread (SIMT) vectorized constant-pressure (\conp/)\slash constant-volume (\conv/) thermo-chemical source-term and sparse\slash dense chemical kinetic Jacobian evaluation has been developed.
Selected chemical kinetic models covering the range of sizes typically used in reactive-flow simulations were used for demonstration.
A new formulation of the chemical kinetic governing equations was derived and validated, resulting Jacobian sparsities of \SIrange{28.6}{92.0}{$\percent$} for the tested models.
Speedups of \SIrange{3.40}{4.08}{$\times$} were found for shallow-vectorized OpenCL source-rate evaluation compared a parallel OpenMP code on an \avx/ central processing unit (CPU), increasing to \SIrange{6.63}{9.44}{$\times$} and \SIrange{3.03}{4.23}{$\times$} for sparse and dense chemical kinetic Jacobian evaluation, respectively.
In addition, the performance of a NVIDIA \gpunew/ graphics processing unit (GPU) was compared to a NVIDIA \gpuold/ GPU finding speedups of \SIrange{1.40}{1.88}{$\times$}, \SIrange{1.10}{1.59}{$\times$}, and \SIrange{1.36}{3.0}{$\times$} for source-rate, sparse and dense Jacobian evaluation, respectively.
Furthermore, the effect of data-ordering was investigated and an storage pattern specifically formulated for vectorized evaluation was proposed; as well, the effect of the \conp/\slash\conv/ assumptions and varying vector widths were studied on source-term evaluation performance.
Speedups over a first-order finite-difference approach reached \SIrange{17.60}{45.13}{$\times$} and \SIrange{55.11}{245.63}{$\times$} for sparse\slash dense evaluation on the CPU\slash GPU respectively, while dense Jacobian evaluation was up to \SI{19.56}{$\times$} and \SI{2.84}{$\times$} times faster than a previous version of \texttt{pyJac} on a \sse/ CPU and \gpuold/ GPU, respectively.
Finally, future directions for vectorized chemical kinetic evaluation and sparse linear-algebra techniques were discussed.
\end{abstract}

@@

As the importance of detailed chemical kinetics for predictive reactive-flow simulations has become recognized~\cite{LU2009192}, the size and complexity of chemical kinetic models describing current and next-generation fuels relevant to transportation and power generation have increased greatly; e.g., a biodiesel model~\cite{WESTBROOK2011742} consists of \textasciitilde\num{3500} chemical species, and over \num{17000} reactions.
Moreover, the cost of evaluating the chemical source-terms scales linearly with the size of the model, while evaluation and factorization of the chemical kinetic Jacobian---if naively implemented via finite-difference method---scale quadratically and cubically with the number of species in the model~\cite{LU2009192}.
These factors often make the use of detailed chemical kinetics prohibitive in practice; e.g., in a direct numerical simulation~\cite{Spafford:2010aa} utilizing a 22-species model, evaluation of reaction rates consumed around half of the total run time.
In addition, chemical kinetic Jacobian evaluation and factorization---utilized by most common implicit integration techniques to deal with numerical stiffness---is a bottleneck for even modestly-sized chemical models in realistic reactive flow simulations, necessitating use of other cost-reduction strategies~\cite{LU2009192}.

A host of techniques have been developed to lessen the computational demand of chemical kinetic calculations while maintaining fidelity, these fall broadly into a few categories: removal of unimportant species and reactions~\cite{Lu:2006bb,Pepiot-Desjardins:2008,Hiremath:2010jw,Niemeyer:2010bt,Curtis:2015}, lumping of species with similar thermo-chemical properties~\cite{Lu:2007,Ahmed:2007fa,Pepiot:2008kq}, and time-scale methods that reduce numerical stiffness~\cite{Maas:1992ws,Lam:1994ws,Lu:2001ve,Gou:2010}.
The interested reader is referred to the recent review of Tur\'anyi and Tomlin~\cite{turanyi2016analysis} for a comprehensive overview.

@@

SIMD and SIMT programming are two important vector-processing paradigms used increasingly in scientific computing.
Traditional multi-core parallelism is often used to increase central processing unit (CPU) performance, however as exponential growth in processing power---colloquially known as Moore's law---has slowed~\cite{khan2018science}, SIMD processors as well as SIMT processors, e.g., in the form of graphics processing units (GPUs), have gained recognition due to their increased floating operation throughput.
The parallel programming standard OpenCL~\cite{stone2010opencl} has further enabled adoption of vector-processing based codes in scientific computing by providing a common application program interface (API) for execution on heterogeneous systems, e.g., CPU, GPU, or Intel's Many Integrated Core (MIC) architecture, etc.
In this discussion, we will largely use OpenCL terminology to describe these processing paradigms, as it provides a convenient way to classify otherwise disparate processor type (e.g., CPUs and GPUs), however the concepts discussed herein are broadly applicable to SIMD\slash SIMT processing and not tied to OpenCL specifically.

\begin{figure}[htb]
  \centering
  \begin{subfigure}[t]{0.45\linewidth}
      \includegraphics[width=\textwidth]{SIMD.pdf}
      \caption{Schematic of SIMD processing.  A single compute unit (e.g., a CPU core) contains a vector unit with $N_v$ processing elements (PEs), together called a vector-lane.  The vector unit executes a single instruction concurrently on multiple data.}
      \label{F:SIMD}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.45\linewidth}
      \includegraphics[width=\textwidth]{SIMT.pdf}
      \caption{Schematic of SIMT processing. A single compute unit (e.g., a GPU streaming multiprocessor) contains many processing elements (PEs) and hosts many threads, each with an instruction to execute (I1, I2).  Threads with the same instruction execute concurrently on multiple data while the others must wait (leading to thread divergence).}
      \label{F:SIMT}
  \end{subfigure}
  \caption{Simple diagrams explaining the fundamentals of the SIMD and SIMT vector-processing paradigms.}
\end{figure}

A typical modern CPU has many compute units (i.e., cores), each with specialized vector processing units capable of running SIMD instructions (\cref{F:SIMD}).
A SIMD instruction utilizes the vector processor to execute the same floating point operation (e.g., multiplication, division, etc.) on a several different data concurrently; the number of concurrent operations possible is known as the vector-width\footnote{OpenCL allows for use of vector-widths different from the actual hardware vector-width via implicit conversion, and may provide some performance benefit as will be studied in Sec.~\ref{S:results}} and is typically around 2--4 double precision operations. 
Additionally, specialized hardware accelerators, e.g., Intel's Xeon Phi co-processor (MIC architecture) have been developed that have tens of cores with very wide vector-widths (e.g., \numrange{4}{8} double precision operations); these very wide vector-widths are also found on cutting-edge and forthcoming Intel CPUs; i.e., the Skylake Xeon and Cannon Lake architectures, respectively.

SIMT processing---the foundation of modern GPUs~\cite{lindholm2008nvidia}---is a related computing paradigm that hosts large numbers of threads on a single compute element (a streaming multiprocessor in NVIDIA terminology).
A group of threads---typically \num{32}, known as a warp on NVIDIA GPUs---execute the same SIMD instruction on multiple data concurrently (\cref{F:SIMT}).
If some threads must execute a different instruction---e.g., due to if\slash then branching or predication---they are forced to wait and execute later.
This phenomenon, known as thread-divergence, is a key consideration for SIMT-processing and can cause serious performance degradation for complicated algorithms~\cite{CurtisGPU:2017}.

@@

Recognizing the need to accelerate Jacobian evaluation and factorization, a number of recent works have been published on analytical chemical kinetic Jacobian evaluation; although as will be discussed at the end of this section, this work offers several key improvements over past efforts.
The \texttt{TChem} software package~\cite{Safta:2011vn} was one of the first packages developed providing this functionality, but has several limitations including: incompatibilities with modern reaction types---i.e., pressure dependent Arrhenius (or P-Log) and Chebyshev reactions---and lack of thread-safety to enable parallel execution~\cite{Curtis2017:tchem}. 
Youssefi~\cite{Youssefi:2011tm} explored the importance of analytical Jacobian matrices for time-scale analysis techniques as well as their effect on computational efficiency in zero-dimensional homogeneous reactor simulations.
Bisetti~\cite{Bisetti:2012jw} developed an isothermal, isobaric analytical Jacobian code-generation utility; although this set of assumptions (particularly, the choice of an isothermal system) are incompatible with many combustion simulations, a significant increase in Jacobian sparsity was exhibited.
Additionally, in the same work Bisetti provided a novel way to compute dense matrix-vector multiplications resulting from a change of system variables without need for storage of the full dense Jacobian.
Perini et al.~\cite{Perini:2012gy} developed an analytical Jacobian code for constant-volume combustion, with additional options to increase sparsity (at expense of strict correctness), and enable tabulation of temperature dependent properties; a \SI{80}{\percent} speedup over a finite-difference-based Jacobian was reported when used in a multidimensional reactive-flow simulation.
Dijkmans et al.~\cite{Dijkmans:2014bb} developed a GPU-based analytical Jacobian code with optional tabulation of temperature-dependent properties, and found speedups up to \SI{120}{$\times$} for zero-dimensional chemical kinetic integration with large chemical models (\textasciitilde\num{3000} species) using a Tesla C2075 GPU.
Niemeyer et al.~\cite{Niemeyer:2016aa} created and validated the open-source analytical chemical kinetic Jacobian code-generator, \texttt{pyJac}, supporting parallel execution on the CPU and SIMT execution on the GPU; a speedup of \SIrange{3}{7.5}{$\times$} over a finite-difference Jacobian was found on the CPU.
Gao et al.~\cite{GAO2015287} derived a sparse analytical Jacobian, but it was not validated outside the context of use with an implicit-integration technique; in addition, since the Jacobian was based on an over-constrained system, the effect on strict conservation of mass\slash energy was not studied.

@@

Niemeyer and Sung~\cite{Niemeyer:2014aa} later developed a stabilized explicit second-order Runge--Kutta--Chebyshev algorithm on a Tesla C2075, demonstrating an order of magnitude speedup for a CPU implementation of VODE on a six-core Intel X5650 for moderately stiff chemical kinetics.

@@

In evaluation of the thermo-chemical properties, speedups between \SIrange{8}{13}{$\times$} were found (over a serial version of the same code) while running on 16 CPU cores, and \SIrange{20}{40}{$\times$} on the GPU; speedups of \textasciitilde\SI{9}{$\times$} and \textasciitilde\SI{25}{$\times$} were found in solution of a partially premixed methanol flame PDE on 16 CPU cores and the GPU, respectively.
Curtis et al.~\cite{CurtisGPU:2017} implemented a fifth-order implicit Runge--Kutta method~\cite{wanner1991solving}, as well as two fourth-order exponential integration techniques~\cite{Hochbruck:1998,Hockbruck:2009} paired with an analytical Jacobian code~\cite{Niemeyer:2016aa} on a Tesla C2075; the implicit Runge--Kutta method was roughly equivalent to a standard implicit integrator~\cite{Hindmarsh:2005} running on 12--38 Intel Xeon E5-4640 v2 CPU cores for two relatively small chemical models with an integration time step of \SI{e-6}{$\sec$}.

In contrast, SIMD-based chemical kinetics evaluation\slash integration have been far less studied.
Linford et al.~\cite{Linford:2011} implemented a three-stage, second-order Rosenbrock integrator for atmospheric chemical kinetics on the CPU, GPU and cell croadband engine (CBE)---a specially designed vector processor---and found speedups regularly exceeding~\SI{25}{$\times$} over a serial CPU implementation.
Kroshko and Spiteri~\cite{kroshko2013efficient} implemented a SIMD-vectorized third-order stiff Rosenbrock integrator for atmospheric chemistry on the CBE and found a speedup of \SI{1.89}{$\times$} (a parallel scaling efficiency of \SI{94}{$\percent$}) over a serial version of the same code.
Stone and Niemeyer~\cite{stone2016} implemented a linearly-implicit fourth-order stiff Rosenbrock solver in the OpenCL for various platforms including CPUs, GPUs, and MICs.
SIMD-vectorization improved the integrator performance over an OpenMP baseline (which was vectorized by simple compiler hints (i.e., \texttt{\#pragmas}) by \SIrange{2.5}{2.8}{$\times$} on the CPU and \SIrange{4.7}{4.9}{$\times$} on the MIC, while the GPU performance was only \SIrange{1.4}{1.6}{$\times$} faster than the OpenMP baseline due to thread-divergence concerns~\cite{stone2016}.

This work will build upon our previous analytical chemical kinetic Jacobian code, \texttt{pyJac}~\cite{Niemeyer:2016aa}, in order to:
\begin{itemize}
 \item Derive and validate a new Jacobian formulation for \texttt{pyJac} that greatly increases sparsity
 \item Enable cross-platform SIMD\slash SIMT vectorization for the CPU, GPU and other accelerators
 \item Investigate the performance for a wide range chemical kinetic models, specifically looking at speedups due to SIMD\slash SIMT-vectorization as well as relative to the previous version of \texttt{pyJac}, and finally
 \item Discuss future extensions to this work as well as several promising directions for SIMD\slash SIMT vectorization in reactive-flow simulations.
\end{itemize}
To our knowledge, this is the first effort that enables vectorized chemical-kinetic source-term and Jacobian evaluation for any chemical model on a wide-selection of platforms, i.e., the CPU, GPU and other accelerators.

@@

\begin{figure}[htb]
  \centering
  \begin{minipage}{0.45\linewidth}
    \begin{subfigure}[t]{\textwidth}
      \includegraphics[width=\textwidth]{data_layouts.pdf}
      \caption{A simple 2-D data array with K rows and N columns.}
      \label{F:mem}
    \end{subfigure}
  \end{minipage}
  \hfil
  \begin{minipage}{0.45\linewidth}
    \begin{subfigure}[t]{\textwidth}
	\includegraphics[width=\textwidth]{row_major.pdf}
	\caption{Row-major data ordering}
	\label{F:row_major}
    \end{subfigure}
    \\
    \\
    \\
    \\
    \begin{subfigure}[t]{\textwidth}
	\includegraphics[width=\textwidth]{column_major.pdf}
	\caption{Column-major data ordering}
	\label{F:column_major}
    \end{subfigure}
  \end{minipage}
\end{figure}

When storing arrays for a chemical kinetic model, the data-storage layout and vectorization patterns are critical to achieving high-performance code.
\Cref{F:mem} depicts an example data array with $K$ rows and $N$ columns, and index ($i$, $j$) corresponding to the $i$th row and $j$th column.
For example, the concentration of species $j$ for the $i$th thermo-chemical state would be stored in $[C]_{i, j}$ with $1 \le i \le N_{state}$ (the number of thermo-chemical states considered for evaluation), and $1 \le j \le \ns/$ (the number of species in the model).
The stored concentrations would then have $K = N_{\text{state}}$ rows, and $N = N_{\text{sp}}$ columns.

In the ``C'' (C row-major) format, the concentrations of all species for a single thermo-chemical condition $i$ would be stored sequentially in memory, i.e., with  $[C]_{1, 1}$ in index \num{1}, $[C]_{1, 2}$ in index \num{2}, etc., as shown in \cref{F:row_major}.
Conversely in the ``F'' (Fortran column-major) format, the concentrations of a single species $j$ over all thermo-chemical states are adjacent in memory, corresponding to storing $[C]_{1, 1}$ in index \num{1}, $[C]_{2, 1}$ in index \num{2} and so on, as shown in \cref{F:column_major}.
This ordering---along with the device (CPU, GPU, etc.\@) and vectorization pattern in question---have a large effect on the performance of SIMD\slash SIMT-vectorized algorithms.

@@

In a \textit{deep}-SIMD\slash SIMT vectorization (also referred to as ``per-block'' in previous GPU works (e.g.,~\cite{Stone:2013aa,CurtisGPU:2017}), a compute-unit utilizes its SIMD-lanes\slash SIMT-threads cooperatively to evaluate the thermo-chemical source-terms for a single thermo-chemical state; thus SIMD-lanes loading $[C]_{1, j} \ldots [C]_{1, j + N_v}$---the species concentrations for state 1, species $j \ldots j + N_v$---will access sequential memory locations if the data is stored in ``C''-order.
Further, in ``C''-ordering the maximum distance between any two species concentrations within the same thermo-chemical state is $N_{\text{sp}}$, with $N_{\text{sp}} \ll N_{\text{state}}$ in most cases; this greatly improved data-locality increases the chances of a cache-hit on the CPU, but may lead to uncoalesced memory-accesses on the GPU.
In addition, a deep vectorization requires both synchronization between SIMD-lanes\slash SIMT-threads via memory-fences\slash barriers (an expensive operation), and may result in SIMD-waste\slash SIMT thread-divergence, caused by different SIMD-lanes\slash SIMT-threads executing different instructions (e.g., resulting from different if\slash then branches).
It is noted that generally speaking, shallow-vectorizations may also experience SIMD-waste\slash SIMT thread-divergence, e.g., in chemical kinetic integration due to varying internal solver time-step sizes~\cite{CurtisGPU:2017}; however, in this work shallow-vectorizations are largely unaffected by this concern.

\begin{figure}[htb]
  \centering
  \begin{minipage}{0.6\linewidth}
    \begin{subfigure}[t]{\textwidth}
	\includegraphics[width=\textwidth]{row_major_split.pdf}
	\caption{Row-major, shallow-vectorized data ordering}
	\label{F:row_major_split}
    \end{subfigure}
    \\
    \\
    \\
    \\
    \begin{subfigure}[t]{\textwidth}
	\includegraphics[width=\textwidth]{column_major_split.pdf}
	\caption{Column-major, deep-vectorized data ordering}
	\label{F:column_major_split}
    \end{subfigure}
  \end{minipage}
  \caption{Vectorized data ordering patterns}
  \label{F:vector_data}
\end{figure}

@@

\subsection{Thermo-chemical Source-Terms and Jacobian}
This new version of \texttt{pyJac} is capable of evaluating the thermo-chemical source-terms for using both the constant-pressure (\conp/) or constant-volume (\conv/) assumption\footnote{Note: in this context, the ``constant-pressure'' and ``constant-volume'' assumptions refer to evaluation within a reaction sub-step in the operator splitting scheme, rather than a general constant-pressure or constant-volume reactive-flow simulation.}
In this section, we will only outline a brief summary of the system evaluated by \texttt{pyJac}; the interested reader is referred to the supplemental material for the complete derivations.
